Toward \earlywarning\ detection of gravitational waves from compact binary coalescence

### 2012-04-04 article.tex
Fixed redline comments
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1510 ccdd05d5-56c1-461a-bf98-85f6876523b4

\usepackage[colorlinks]{hyperref}
\usepackage{epstopdf}
\usepackage{amsmath,amsthm}
\usepackage{stmaryrd} % For \shortuparrow and \shortdownarrow
\usepackage{upgreek} % For \upmu, upright mu

@@

\newcommand{\earlywarning}{early-warning}
\newcommand{\NS}{NS}
\newcommand{\mbta}{MBTA}
\newcommand{\GW}{GW}%{{\sc gw}}
\newcommand{\EM}{EM}%{{\sc em}}
\newcommand{\GRB}{GRB}%{{\sc grb}}
\newcommand{\CBC}{CBC}%{{\sc cbc}}
\newcommand{\LIGO}{LIGO}%{{\sc ligo}}
\newcommand{\LCGT}{LCGT}%{{\sc lcgt}}
\newcommand{\GEO}{GEO600}%{{\sc geo600}}
\newcommand{\LSO}{LSO}%{{\sc lso}}
\newcommand{\SNR}{SNR}%{{\sc snr}}
\newcommand{\realtime}{real-time}
\newcommand{\Msun}{\ensuremath{M_{\odot}}}
\newcommand{\order}[1]{\ensuremath{\mathcal{O}[#1]}}
\newcommand{\tmpsamps}{\ensuremath{N}}
\newcommand{\numtmps}{\ensuremath{M}}
\newcommand{\numslices}{\ensuremath{S}}
% macros for number of svd basis functions
\newcommand{\SVD}{SVD}%{{\sc svd}}
\newcommand{\svdtmps}[1]{\ensuremath{L^#1}}
\newcommand{\numsvdtmps}{\svdtmps{s}}
% macros for sample points in slices
\newcommand{\slicesamps}[1]{\ensuremath{N^#1}}
\newcommand{\slicessamps}{\slicesamps{s}}
\newcommand{\fftblock}{\ensuremath{D}}
\newcommand{\resampsamps}{\ensuremath{N^\shortdownarrow,\, N^\shortuparrow}}
\newcommand{\fir}{FIR}%{{\sc fir}}
\newcommand{\fft}{FFT}%{{\sc fft}}
\newcommand{\fmax}{\ensuremath{f^0}}
\newcommand{\flops}{flop/s}
\newcommand{\gstlal}{{\tt gstlal}}
\newcommand{\gstreamer}{GStreamer}
\newcommand{\numcpus}{{600}}
\newcommand{\lloid}{LLOID}%{{\sc lloid}}
\newcommand{\TD}{TD}%{{\sc td}}
\newcommand{\FD}{FD}%{{\sc fd}}

@@

\keywords{gamma ray burst: general --- gravitational waves --- methods: data analysis --- methods: numerical}

\begin{abstract}
Rapid detection of compact binary coalescence with a network of advanced
gravitational-wave detectors will offer a unique opportunity for
multi-messenger astronomy.  Prompt detection alerts for the astronomical
community might make it possible to observe the onset of electromagnetic
emission from compact binary coalescence.  We demonstrate a computationally
practical filtering strategy that could produce early-warning triggers before
gravitational radiation from the final merger has arrived at the detectors.
\end{abstract}

@@

As a compact binary system loses energy to gravitational waves (\GW{}s), its
orbital separation decays, leading to a runaway inspiral with the \GW\
amplitude and frequency increasing until the system eventually merges.  If a
neutron star (\NS) is involved, it might become tidally disrupted near the
merger and fuel an electromagnetic (\EM) counterpart \citep{shibata:2007}.
Effort from both the \GW\ and the broader astronomical communities might make
it possible to use \GW\ observations as early warning triggers for \EM\
followup. In the first generation of ground-based laser interferometers, the
\GW\ community initiated a project to send alerts when potential \GW\
transients were observed in order to trigger followup observations by \EM\
telescopes.  The typical latencies were 30 minutes \citep{HugheyGWPAW2011}.
This was an important achievement, but too late to catch any prompt optical flash
or the onset of an on-axis optical afterglow.  Since the \GW\ signal is in principle detectable even before the tidal
disruption, one might have the ambition of reporting \GW\ candidates not minutes
after the merger, but seconds before.  We explore one essential ingredient of
this problem, a computationally inexpensive latency free, real-time filtering
algorithm for detecting inspiral signals in \GW\ data.  We also consider the
prospects for advanced \GW\ detectors and discuss other areas of work that would
be required for rapid analysis.

Compact binary coalescence (\CBC) is a plausible progenitor for most short
gamma-ray bursts (short \GRB{}s) \citep{Lee:2005, nakar07}, but the
association is not iron-clad \citep{2011ApJ...727..109V}. The tidally
disrupted material falls onto the newly formed, rapidly spinning compact object
and is accelerated in jets along the spin axis with a timescale of $0.1$--$1$~s
after the merger \citep{Janka1999}, matching the short \GRB\ duration
distribution well. Prompt \EM\ emission including the \GRB\ can arise as fast
outflowing matter collides with slower matter ejected earlier in inner shocks.
The same inner shocks, or potentially reverse shocks, can produce an
accompanying optical flash \citep{Sari99}. The prompt emission is a probe into
the extreme initial conditions of the outflow, in contrast with afterglows,
which arise in the external shock with the local medium and are relatively
insensitive to initial conditions. Optical flashes have been observed for a
handful of long \GRB{}s \citep{2011CRPhy..12..255A} by telescopes with extremely
rapid response or, in the case of GRB 080319b, by pure serendipity, where
several telescopes were already observing the afterglow of another \GRB\ in
the same field of view \citep{2008Natur.455..183R}. The observed optical flashes
peaked within tens of seconds and decayed quickly. For short \GRB\ energy
balance and plasma density, however, the reverse shock model predicts a peak
flux in radio, approximately 20 minutes after the \GRB, but also a relatively
faint optical flash \citep{nakar07}; for a once-per-year Advanced LIGO event at
$130$~Mpc, the radio flux will peak around 9~GHz at $\sim$$5$~mJy, with emission
in the $R$-band at $\sim$19~mag. Interestingly, roughly a quarter to half of
observed short \GRB{}s also exhibit extended X-ray emission of $30$--$100$~s in
duration beginning $\sim$$10$~s after the \GRB\ and carrying comparable fluence
to the initial outburst. This can be explained if the merger results in the
formation of a proto-magnetar that interacts with
ejecta~\citep{Bucciantini2011}. Rapid \GW\ alerts would enable joint \EM\ and
\GW\ observations to confirm the short \GRB-\CBC\ link and allow the early
\EM\ observation of exceptionally nearby and thus bright events.

In October 2010, \LIGO{}\footnote{\url{http://www.ligo.org/}} completed its
sixth science run (S6) and Virgo\footnote{\url{http://www.ego-gw.it/}}
completed its third science run (VSR3).  While both \LIGO\ detectors and Virgo
were operating, several all-sky detection pipelines operated in a low-latency
configuration to send astronomical alerts, namely Coherent WaveBurst (cWB), Omega, and Multi-Band Template Analysis \citep[MBTA;][]{HugheyGWPAW2011, S6lowlatency2, S6lowlatency3, S6lowlatency4}.  cWB and Omega are both unmodeled searches for bursts based on time-frequency decomposition of the GW data.  \mbta\ is a novel kind of template-based inspiral search that was purpose-built for low latency operation.  \mbta\ achieved the best \GW\ trigger-generation latencies of 2--5 minutes.
Alerts were sent with latencies of 30--60 minutes, dominated by human vetting.
Candidates were sent for \EM\ followup to several telescopes; Swift,
LOFAR, ROTSE, TAROT, QUEST, SkyMapper,
Liverpool Telescope, Pi of the Sky, Zadko, and Palomar Transient Factory
\citep{kanner2008, HugheyGWPAW2011} imaged some of the most likely sky
locations.

@@

\paragraph{Data conditioning ($\sim$1~min)}%
Science data must be calibrated using the detector's frequency response to
gravitational radiation.  Currently, data are calibrated in blocks of 16~s.
Within $\sim$1~min, data quality is assessed in order to create veto flags.
These are both technical sources of latency that might be addressed with
improved calibration and data quality software for advanced detectors.

\paragraph{Trigger generation (2--5~min)}%
Low-latency data analysis pipelines
deployed in S6/VSR3 achieved an impressive latency of minutes.  However, second
to the human vetting process, this dominated the latency of the entire \EM\
followup process.  Even if no other sources of latency existed, this trigger
generation latency is too long to catch prompt or even extended emission.
Low-latency trigger generation will become more challenging with advanced
detectors because inspiral signals will stay in band up to ten times longer.  In
this work, we will focus on reducing this source of latency.

@@

This work will focus on reducing the latency of trigger production.  Data
analysis strategies for advance detection of \CBC{}s will have to strike a
balance between latency and throughput. \CBC\ searches consist of banks of
matched filters, or cross-correlations between the data stream and a bank of
nominal ``template'' signals.  There are many different implementations of
matched filters, but most have high throughput at the cost of high latency, or
low latency at the cost of low throughput.  The former are epitomized by the
overlap-save algorithm for frequency domain (\FD) convolution, currently the
preferred method in \GW\ searches.  The most obvious example of the latter is
direct time domain (\TD) convolution, which is latency-free.  However, its cost in floating
point operations per second is linear in the length of the templates, so it is
prohibitively expensive for long templates.  The computational challenges of low-latency \CBC\ searches are still more daunting for advanced detectors for which the inspiral signal remains in band for a large fraction of an hour~(see appendix~\ref{sec:low-frequency-cutoff}).

@@

Several efforts that exploit one or both of these properties are under way to
develop low-latency \CBC\ search pipelines with tractable computing requirements.
One example is \mbta{}~\citep{Marion2004, Buskulic2010}, which was deployed in
S6/VSR3.  MBTA consists of multiple, usually two, template banks for different frequency bands, one which is matched to the early inspiral and the other which is matched to the late inspiral.  An excursion in the output of any filter bank triggers coherent reconstruction of the full matched filtered output.  Final triggers are built from the reconstructed matched filter output.  Another novel approach using networks of parallel, second order infinite impulse response~(IIR) filters is being explored
by \citet{shaunIIR, linqingIIR}.

@@

The paper is organized as follows.  First, we discuss prospects for
early-warning detection.  Then, we provide an overview of our novel
method for detecting compact binary coalescence signals near real-time.
We then describe a prototype implementation using open source signal
processing software.  To validate our approach we present a case study
focusing on a particular subset of the \NS--\NS\ parameter space.  We
conclude with some remarks on what remains to prepare for the advanced
detector era.

\section{Prospects for \earlywarning\ detection and \EM\ followup}

\begin{figure}[h]
\plotone{f1}
\caption{\label{fig:earlywarning}Expected number of \NS--\NS\
sources that could be detectable by Advanced \LIGO\ a given number of seconds
before coalescence.  The heavy solid line corresponds to the most probable yearly rate
estimate from~\citet{Abadie:2010p10836}.  The shaded region represents the 5\% to 95\% confidence interval
arising from substantial uncertainty in predicted event rates.}
\end{figure}
%
Before the \GW\ signal leaves the detection band, we can imagine examining the
signal to noise ratio (\SNR) accumulated up to that point and if it is
already significant, release an alert immediately, trading \SNR\ and sky
localization accuracy for pre-merger detection.

In the quadrupole approximation, the instantaneous frequency of the GW inspiral signal is related to the time $t$ relative to coalescence ~\citep[section 5.1 of][]{livrev12} through
%
\begin{equation} \label{eq:fgw}
	f(t) = \frac{1}{\pi \mathcal{M}_\mathrm{t}}
		\left[ \frac{5}{256}\frac{\mathcal{M}_\mathrm{t}}{t} \right]^{3/8}
\end{equation}
%
where $\mathcal{M}=M^{2/5} \mu^{3/5}$ is the chirp mass of the binary, $\mathcal{M}_\mathrm{t}=G \mathcal{M} / c^3$ is the chirp mass in units of time, $M$ is the total mass, and $\mu$ is the reduced mass. 
The expected value of the single detector \SNR\ for an optimally oriented (source at detector's zenith or nadir, orbital plane face-on) inspiral source is~\citep{Abadie:2010p10836}
%
\begin{equation}
	\label{eq:expected-snr}
	\rho =
		\frac{{\mathcal{M}_\mathrm{t}}^{5/6} c}{\pi^{2/3} D}
		\sqrt{
			\frac{5}{6} \int_{f_\mathrm{low}}^{f_\mathrm{high}}
			\frac{f^{-7/3}}{S(f)} \mathrm{d}f}
\end{equation}
%
where $D$ is the luminosity distance and $S(f)$ is the one-sided power spectral density of the detector noise.  $f_\mathrm{low}$ and $f_\mathrm{high}$ are low and high frequency limits of integration which may be chosen to extend across the entire bandwidth of the detector.  If we want to trigger at a time $t$ before merger, then we must cut off the SNR integration at $f_\mathrm{high} = f(t)$ with $f(t)$ given by equation~(\ref{eq:fgw}) above.

Figure~\ref{fig:earlywarning} shows projected early detectability rates for
\NS--\NS\ binaries in Advanced \LIGO\ assuming the anticipated detector sensitivity for
the `zero detuning, high power' configuration described in \citet{ALIGONoise} and
\NS--\NS\ merger rates estimated in \citet{Abadie:2010p10836}.  The merger rates
have substantial measurement uncertainty due to the small sample of known double pulsar systems that will merge within a Hubble time; they also have systematic uncertainty due sensitive dependence on the pulsar luminosity distribution function~\citep{KalogeraRates}.
The most probable estimates indicate that at a single-detector \SNR\ threshold of 8, we will
observe a total of 40~events~yr$^{-1}$; $\sim$10~yr$^{-1}$ will be detectable
within 10~s of merger and $\sim$5~yr$^{-1}$ will be detectable within 25~s of
merger if analysis can proceed with near zero latency.

@@

\begin{table}[h]
\caption{\label{table:sky-localization-accuracy}Horizon distance, \SNR\ at
merger, and area of 90\% confidence at selected times before merger for sources
with expected detectability rates of 40, 10, 1, and 0.1~yr$^{-1}$.  A dash (-----) signifies that the confidence area is omitted because at the indicated time the SNR would not have crossed the detection threshold of 8.}
\begin{center}
\begin{tabular}{rrrrrrr}
\tableline\tableline
rate & horiz. & final & \multicolumn{4}{c}{$A$(90\%) (deg$^2$)} \\
\cline{4-7}
yr$^{-1}$ & (Mpc) & \SNR\ & 25 s & 10 s & 1 s & 0 s \\
\tableline
40\phd\phn & 445 & 8.0 & ----- & ----- & ----- & 9.6 \\
10\phd\phn & 280 & 12.7 & ----- & 1200 & 78 & 3.8 \\
1\phd\phn & 130 & 27.4 & 1300 & 260 & 17 & 0.8 \\
0.1 & 60 & 58.9 & 280 & 56 & 3.6 & 0.2 \\
\tableline
\end{tabular}
\end{center}
\end{table}
%
\begin{figure}[h]
\plotone{f2}
\caption{\label{fig:sky-localization-accuracy}Area of the 90\% confidence
region as a function of time before coalescence for sources with anticipated
detectability rates of 40, 10, 1, and 0.1~yr$^{-1}$. The heavy dot indicates
the time at which the accumulated \SNR\ exceeds a single-detector threshold of~8.}
\end{figure}

\EM\ followup requires estimating the location of the \GW\ source. The localization
uncertainty can be estimated from the uncertainty in the time of arrival of the \GW{}s,
which is determined by the signal's effective bandwidth and \SNR\
\citep{Fairhurst2009}.  Table~\ref{fig:earlywarning} and
Figure~\ref{fig:sky-localization-accuracy} show the estimated 90\%
confidence area versus time of the loudest coalescence events detectable by
Advanced \LIGO\ and Advanced Virgo.  This is the \emph{minimum} area; localization is best at high elevation
from the plane containing the detectors and worst at zero elevation.  \citeauthor{Fairhurst2009} also cautions that his Fisher matrix calculation fails to capture disconnected patches of probability, which occur prominently in networks of three detectors where there are generally two local maxima on opposite sides of the plane of the detectors.  Aside from the mirror degeneracy, characterizing the uncertainty region by the Fisher matrix alone tends to overestimate, rather than underestimate, the area for low SNR events, but this effect is generally more than compensated by the source being in an unfavorable sky location.  For these reasons, the localization uncertainty estimated from timing is highly optimistic and will only suffice for an order of magnitude estimate.  Once per year, we expect to observe an
event with a final single-detector \SNR\ of $\approx$27 whose location can be constrained to about
1300~deg$^2$ (3.1\% of the sky) within 25~s of merger,
260~deg$^2$ (0.63\% of the sky) within 10~s of merger, and
0.82~deg$^2$ (0.0020\% of the sky) at merger.

It is unfeasible to search hundreds of square degrees for a prompt counterpart.  For comparison to some examples of modern ground-based wide-field survey instruments, the Palomar Transient Factory P48 ~\citep{2010SPIE.7735E.122L} has a $3.50 \times 2.31$~deg$^2$ field of view~(FOV); the Pan-STARRS P1~\citep{2002SPIE.4836..154K} has a 7~deg$^2$ FOV.  Even the eagerly awaited LSST will have a FOV of 9.6~deg$^2$ \citep{2008arXiv0805.2366I}.  However, 
it is possible to reduce the localization uncertainty by only looking at
galaxies from a catalog that lie near the sky location and luminosity distance
estimate from the \GW\ signal~\citep{galaxy-catalog} as was done in S6/VSR3.
Within the expected Advanced \LIGO\ \NS--\NS\ horizon distance,
the number of galaxies that can produce a given signal amplitude is much larger
than in Initial \LIGO\ and thus the catalog will not be as useful
for downselecting pointings for most events. However, exceptional \GW\ sources will
necessarily be extremely nearby. Within this reduced volume there will be fewer
galaxies to consider for a given candidate and catalog completeness will be
less of a concern.  This should reduce the 90\% confidence area substantially.

@@

In the next section we describe how we form our early-warning detection
statistic using the time slice decomposition and the \SVD.

>>>>

\usepackage[colorlinks]{hyperref}
\usepackage{epstopdf}
\usepackage{amsmath,amsthm}
\usepackage{stmaryrd} % For \shortuparrow and \shortdownarrow
\usepackage{upgreek} % For \upmu, upright mu
\usepackage{multirow}

@@

\newcommand{\earlywarning}{early-warning}
\newcommand{\NS}{NS}
\newcommand{\mbta}{MBTA}
\newcommand{\GW}{GW}%{{\sc gw}}
\newcommand{\EM}{EM}%{{\sc em}}
\newcommand{\GRB}{GRB}%{{\sc grb}}
\newcommand{\CBC}{CBC}%{{\sc cbc}}
\newcommand{\LIGO}{LIGO}%{{\sc ligo}}
\newcommand{\LCGT}{LCGT}%{{\sc lcgt}}
\newcommand{\GEO}{GEO600}%{{\sc geo600}}
\newcommand{\LSO}{LSO}%{{\sc lso}}
\newcommand{\SNR}{S/N}%{{\sc snr}}
\newcommand{\realtime}{real-time}
\newcommand{\Msun}{\ensuremath{M_{\odot}}}
\newcommand{\order}[1]{\ensuremath{\mathcal{O}[#1]}}
\newcommand{\tmpsamps}{\ensuremath{N}}
\newcommand{\numtmps}{\ensuremath{M}}
\newcommand{\numslices}{\ensuremath{S}}
% macros for number of svd basis functions
\newcommand{\SVD}{SVD}%{{\sc svd}}
\newcommand{\svdtmps}[1]{\ensuremath{L^#1}}
\newcommand{\numsvdtmps}{\svdtmps{s}}
% macros for sample points in slices
\newcommand{\slicesamps}[1]{\ensuremath{N^#1}}
\newcommand{\slicessamps}{\slicesamps{s}}
\newcommand{\fftblock}{\ensuremath{D}}
\newcommand{\resampsamps}{\ensuremath{N^\shortdownarrow,\, N^\shortuparrow}}
\newcommand{\fir}{FIR}%{{\sc fir}}
\newcommand{\fft}{FFT}%{{\sc fft}}
\newcommand{\fmax}{\ensuremath{f^0}}
\newcommand{\flops}{flop~s$^{-1}$}
\newcommand{\gstlal}{{\tt gstlal}}
\newcommand{\gstreamer}{GStreamer}
\newcommand{\numcpus}{{600}}
\newcommand{\lloid}{LLOID}%{{\sc lloid}}
\newcommand{\TD}{TD}%{{\sc td}}
\newcommand{\FD}{FD}%{{\sc fd}}

@@

\keywords{gamma-ray burst: general --- gravitational waves --- methods: data analysis --- methods: numerical}

\begin{abstract}
Rapid detection of compact binary coalescence (CBC) with a network of advanced
gravitational-wave detectors will offer a unique opportunity for
multi-messenger astronomy.  Prompt detection alerts for the astronomical
community might make it possible to observe the onset of electromagnetic
emission from CBC.  We demonstrate a computationally
practical filtering strategy that could produce early-warning triggers before
gravitational radiation from the final merger has arrived at the detectors.
\end{abstract}

@@

As a compact binary system loses energy to gravitational waves (\GW{}s), its
orbital separation decays, leading to a runaway inspiral with the \GW\
amplitude and frequency increasing until the system eventually merges.  If a
neutron star (\NS) is involved, it might become tidally disrupted near the
merger and fuel an electromagnetic (\EM) counterpart \citep{shibata:2007}.
Effort from both the \GW\ and the broader astronomical communities might make
it possible to use \GW\ observations as early warning triggers for \EM\
follow-up. In the first generation of ground-based laser interferometers, the
\GW\ community initiated a project to send alerts when potential \GW\
transients were observed in order to trigger follow-up observations by \EM\
telescopes.  The typical latencies were 30 minutes \citep{HugheyGWPAW2011}.
This was an important achievement, but too late to catch any prompt optical flash
or the onset of an on-axis optical afterglow.  Since the \GW\ signal is in principle detectable even before the tidal
disruption, one might have the ambition of reporting \GW\ candidates not minutes
after the merger, but seconds before.  We explore one essential ingredient of
this problem, a computationally inexpensive latency free, real-time filtering
algorithm for detecting inspiral signals in \GW\ data.  We also consider the
prospects for advanced \GW\ detectors and discuss other areas of work that would
be required for rapid analysis.

Compact binary coalescence (\CBC) is a plausible progenitor for most short
gamma-ray bursts \citep[short \GRB{}s;][]{Lee:2005, nakar07}, but the
association is not iron-clad \citep{2011ApJ...727..109V}. The tidally
disrupted material falls onto the newly formed, rapidly spinning compact object
and is accelerated in jets along the spin axis with a timescale of $0.1$--$1$~s
after the merger \citep{Janka1999}, matching the short \GRB\ duration
distribution well. Prompt \EM\ emission including the \GRB\ can arise as fast
outflowing matter collides with slower matter ejected earlier in inner shocks.
The same inner shocks, or potentially reverse shocks, can produce an
accompanying optical flash \citep{Sari99}. The prompt emission is a probe into
the extreme initial conditions of the outflow, in contrast with afterglows,
which arise in the external shock with the local medium and are relatively
insensitive to initial conditions. Optical flashes have been observed for a
handful of long \GRB{}s \citep{2011CRPhy..12..255A} by telescopes with extremely
rapid response or, in the case of GRB 080319b, by pure serendipity, where
several telescopes were already observing the afterglow of another \GRB\ in
the same field of view \citep[FOV;][]{2008Natur.455..183R}. The observed optical flashes
peaked within tens of seconds and decayed quickly. For short \GRB\ energy
balance and plasma density, however, the reverse shock model predicts a peak
flux in radio, approximately 20 minutes after the \GRB, but also a relatively
faint optical flash \citep{nakar07}; for a once-per-year Advanced LIGO event at
$130$~Mpc, the radio flux will peak around 9~GHz at $\sim$$5$~mJy, with emission
in the $R$-band at $\sim$19~mag. Interestingly, roughly a quarter to half of the 
observed short \GRB{}s also exhibit extended X-ray emission of $30$--$100$~s in
duration beginning $\sim$$10$~s after the \GRB\ and carrying comparable fluence
to the initial outburst. This can be explained if the merger results in the
formation of a proto-magnetar that interacts with
ejecta~\citep{Bucciantini2011}. Rapid \GW\ alerts would enable joint \EM\ and
\GW\ observations to confirm the short \GRB-\CBC\ link and allow the early
\EM\ observation of exceptionally nearby and thus bright events.

In 2010 October, \LIGO{}\footnote{\url{http://www.ligo.org/}} completed its
sixth science run (S6) and Virgo\footnote{\url{http://www.ego-gw.it/}}
completed its third science run (VSR3).  While both \LIGO\ detectors and Virgo
were operating, several all-sky detection pipelines operated in a low-latency
configuration to send astronomical alerts, namely, Coherent WaveBurst (cWB), Omega, and Multi-Band Template Analysis \citep[MBTA;][]{HugheyGWPAW2011, S6lowlatency2, S6lowlatency3, S6lowlatency4}.  cWB and Omega are both unmodeled searches for bursts based on time-frequency decomposition of the GW data.  \mbta\ is a novel kind of template-based inspiral search that was purpose-built for low latency operation.  \mbta\ achieved the best \GW\ trigger-generation latencies of 2--5 minutes.
Alerts were sent with latencies of 30--60 minutes, dominated by human vetting.
Candidates were sent for \EM\ follow-up to several telescopes; \textit{Swift},
LOFAR, ROTSE, TAROT, QUEST, SkyMapper,
Liverpool Telescope, Pi of the Sky, Zadko, and Palomar Transient Factory
\citep{kanner2008, HugheyGWPAW2011} imaged some of the most likely sky
locations.

@@

\paragraph{Data conditioning ($\sim$1~min)}%
Science data must be calibrated using the detector's frequency response to
gravitational radiation.  Currently, data are calibrated in blocks of 16~s.
Within $\sim$1~minute, data quality is assessed in order to create veto flags.
These are both technical sources of latency that might be addressed with
improved calibration and data quality software for advanced detectors.

\paragraph{Trigger generation (2--5~min)}%
Low-latency data analysis pipelines
deployed in S6/VSR3 achieved an impressive latency of minutes.  However, second
to the human vetting process, this dominated the latency of the entire \EM\
follow-up process.  Even if no other sources of latency existed, this trigger
generation latency is too long to catch prompt or even extended emission.
Low-latency trigger generation will become more challenging with advanced
detectors because inspiral signals will stay in band up to 10 times longer.  In
this work, we will focus on reducing this source of latency.

@@

This work will focus on reducing the latency of trigger production.  Data
analysis strategies for advance detection of \CBC{}s will have to strike a
balance between latency and throughput. \CBC\ searches consist of banks of
matched filters, or cross-correlations between the data stream and a bank of
nominal ``template'' signals.  There are many different implementations of
matched filters, but most have high throughput at the cost of high latency, or
low latency at the cost of low throughput.  The former are epitomized by the
overlap-save algorithm for frequency-domain (\FD) convolution, currently the
preferred method in \GW\ searches.  The most obvious example of the latter is
direct time domain (\TD) convolution, which is latency-free.  However, its cost in floating
point operations per second is linear in the length of the templates, so it is
prohibitively expensive for long templates.  The computational challenges of low-latency \CBC\ searches are still more daunting for advanced detectors for which the inspiral signal remains in band for a large fraction of an hour~(see the Appendix).

@@

Several efforts that exploit one or both of these properties are under way to
develop low-latency \CBC\ search pipelines with tractable computing requirements.
One example is \mbta{}~\citep{Marion2004, Buskulic2010}, which was deployed in
S6/VSR3.  MBTA consists of multiple, usually two, template banks for different frequency bands, one which is matched to the early inspiral and the other which is matched to the late inspiral.  An excursion in the output of any filter bank triggers coherent reconstruction of the full matched filtered output.  Final triggers are built from the reconstructed matched filter output.  Another novel approach using networks of parallel, second-order infinite impulse response~(IIR) filters is being explored
by \citet{shaunIIR} and \citet{linqingIIR}.

@@

The paper is organized as follows.  First, we discuss prospects for
early-warning detection.  Then, we provide an overview of our novel
method for detecting CBC signals near real-time.
We then describe a prototype implementation using open source signal
processing software.  To validate our approach we present a case study
focusing on a particular subset of the \NS--\NS\ parameter space.  We
conclude with some remarks on what remains to prepare for the advanced
detector era.

\section{Prospects for \earlywarning\ detection and \EM\ follow-up}

\begin{figure}[h]
\plotone{f1}
\caption{\label{fig:earlywarning}Expected number of \NS--\NS\
sources that could be detectable by Advanced \LIGO\ a given number of seconds
before coalescence.  The heavy solid line corresponds to the most probable yearly rate
estimate from~\citet{Abadie:2010p10836}.  The shaded region represents the 5\%--95\% confidence interval
arising from substantial uncertainty in predicted event rates.}
\end{figure}
%
Before the \GW\ signal leaves the detection band, we can imagine examining the
signal-to-noise ratio (\SNR) accumulated up to that point and if it is
already significant, release an alert immediately, trading \SNR\ and sky
localization accuracy for pre-merger detection.

In the quadrupole approximation, the instantaneous frequency of the GW inspiral signal is related to the time $t$ relative to coalescence ~\citep[section 5.1 of][]{livrev12} through
%
\begin{equation} \label{eq:fgw}
	f(t) = \frac{1}{\pi \mathcal{M}_\mathrm{t}}
		\left[ \frac{5}{256}\frac{\mathcal{M}_\mathrm{t}}{t} \right]^{3/8},
\end{equation}
%
where $\mathcal{M}=M^{2/5} \mu^{3/5}$ is the chirp mass of the binary, $\mathcal{M}_\mathrm{t}=G \mathcal{M} / c^3$ is the chirp mass in units of time, $M$ is the total mass, and $\mu$ is the reduced mass. 
The expected value of the single-detector \SNR\ for an optimally oriented (source at detector's zenith or nadir, orbital plane face-on) inspiral source is~\citep{Abadie:2010p10836}
%
\begin{equation}
	\label{eq:expected-snr}
	\rho =
		\frac{{\mathcal{M}_\mathrm{t}}^{5/6} c}{\pi^{2/3} D}
		\sqrt{
			\frac{5}{6} \int_{f_\mathrm{low}}^{f_\mathrm{high}}
			\frac{f^{-7/3}}{S(f)} \mathrm{d}f},
\end{equation}
%
where $D$ is the luminosity distance and $S(f)$ is the one-sided power spectral density of the detector noise.  $f_\mathrm{low}$ and $f_\mathrm{high}$ are low- and high- frequency limits of integration which may be chosen to extend across the entire bandwidth of the detector.  If we want to trigger at a time $t$ before merger, then we must cut off the SNR integration at $f_\mathrm{high} = f(t)$ with $f(t)$ given by Equation~(\ref{eq:fgw}) above.

Figure~\ref{fig:earlywarning} shows projected early detectability rates for
\NS--\NS\ binaries in Advanced \LIGO\ assuming the anticipated detector sensitivity for
the `zero detuning, high power' configuration described in \citet{ALIGONoise} and
\NS--\NS\ merger rates estimated in \citet{Abadie:2010p10836}.  The merger rates
have substantial measurement uncertainty due to the small sample of known double pulsar systems that will merge within a Hubble time; they also have systematic uncertainty due to sensitive dependence on the pulsar luminosity distribution function~\citep{KalogeraRates}.
The most probable estimates indicate that at a single-detector \SNR\ threshold of 8, we will
observe a total of 40~events~yr$^{-1}$; $\sim$10~yr$^{-1}$ will be detectable
within 10~s of merger and $\sim$5~yr$^{-1}$ will be detectable within 25~s of
merger if analysis can proceed with near zero latency.

@@

\begin{table}[h]
\caption{\label{table:sky-localization-accuracy}Horizon Distance, \SNR\ at
Merger, and Area of 90\% Confidence at Selected Times Before Merger for Sources
with Expected Detectability Rates of 40, 10, 1, and 0.1~yr$^{-1}$.}
\begin{center}
\begin{tabular}{rrrrrrr}
\tableline\tableline
Rate & Horizon & Final & \multicolumn{4}{c}{$A$(90\%) (deg$^2$)} \\
\cline{4-7}
yr$^{-1}$ & (Mpc) & \SNR\ & 25 s & 10 s & 1 s & 0 s \\
\tableline
40\phd\phn & 445 & 8.0 & ----- & ----- & ----- & 9.6 \\
10\phd\phn & 280 & 12.7 & ----- & 1200 & 78 & 3.8 \\
1\phd\phn & 130 & 27.4 & 1300 & 260 & 17 & 0.8 \\
0.1 & 60 & 58.9 & 280 & 56 & 3.6 & 0.2 \\
\tableline
\end{tabular}
\tablecomments{A dash (-----) signifies that the confidence area\\is omitted because at the indicated time the SNR would\\not have crossed the detection threshold of 8.}
\end{center}
\end{table}
%
\begin{figure}[h]
\plotone{f2}
\caption{\label{fig:sky-localization-accuracy}Area of the 90\% confidence
region as a function of time before coalescence for sources with anticipated
detectability rates of 40, 10, 1, and 0.1~yr$^{-1}$. The heavy dot indicates
the time at which the accumulated \SNR\ exceeds a single-detector threshold of~8.}
\end{figure}

\EM\ follow-up requires estimating the location of the \GW\ source. The localization
uncertainty can be estimated from the uncertainty in the time of arrival of the \GW{}s,
which is determined by the signal's effective bandwidth and \SNR\
\citep{Fairhurst2009}.  Table~\ref{fig:earlywarning} and
Figure~\ref{fig:sky-localization-accuracy} show the estimated 90\%
confidence area versus time of the loudest coalescence events detectable by
Advanced \LIGO\ and Advanced Virgo.  This is the \emph{minimum} area; localization is best at high elevation
from the plane containing the detectors and worst at zero elevation.  \citeauthor{Fairhurst2009} also cautions that his Fisher matrix calculation fails to capture disconnected patches of probability, which occur prominently in networks of three detectors where there are generally two local maxima on opposite sides of the plane of the detectors.  Aside from the mirror degeneracy, characterizing the uncertainty region by the Fisher matrix alone tends to overestimate, rather than underestimate, the area for low-\SNR\ events, but this effect is generally more than compensated by the source being in an unfavorable sky location.  For these reasons, the localization uncertainty estimated from timing is highly optimistic and will only suffice for an order-of-magnitude estimate.  Once per year, we expect to observe an
event with a final single-detector \SNR\ of $\approx$27 whose location can be constrained to about
1300~deg$^2$ (3.1\% of the sky) within 25~s of merger,
260~deg$^2$ (0.63\% of the sky) within 10~s of merger, and
0.82~deg$^2$ (0.0020\% of the sky) at merger.

It is unfeasible to search hundreds of square degrees for a prompt counterpart.  For comparison to some examples of modern ground-based wide-field survey instruments, the Palomar Transient Factory P48 ~\citep{2010SPIE.7735E.122L} has a $3.50 \times 2.31$~deg$^2$ FOV; the Pan-STARRS P1~\citep{2002SPIE.4836..154K} has a 7~deg$^2$ FOV.  Even the eagerly awaited LSST will have an FOV of 9.6~deg$^2$ \citep{2008arXiv0805.2366I}.  However, 
it is possible to reduce the localization uncertainty by only looking at
galaxies from a catalog that lie near the sky location and luminosity distance
estimate from the \GW\ signal~\citep{galaxy-catalog} as was done in S6/VSR3.
Within the expected Advanced \LIGO\ \NS--\NS\ horizon distance,
the number of galaxies that can produce a given signal amplitude is much larger
than in Initial \LIGO\ and thus the catalog will not be as useful
for downselecting pointings for most events. However, exceptional \GW\ sources will
necessarily be extremely nearby. Within this reduced volume there will be fewer
galaxies to consider for a given candidate and catalog completeness will be
less of a concern.  This should reduce the 90\% confidence area substantially.

@@

In the next section, we describe how we form our early-warning detection
statistic using the time slice decomposition and the \SVD.
